Type Evolution in Interactive Programming Systems

### 2024-05-31 paper.tex
editing live modeling

\begin{figure}[t]
\begin{adjustwidth}{-1.5em}{-1.5em}
\begin{minipage}[t]{0.26\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class Machine
  name: string
  states: State*
  vars: Decl*
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.28\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class State
  name: string
  onEntry: Stmt*
  transitions: Trans*
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.24\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class Trans
  event: string
  target: State
$~$
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.24\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class Decl
  name: string
  type: Type
$~$
\end{lstlisting}
\end{minipage}
\end{adjustwidth}
\vspace{-1em}
\caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
\label{lst:machinedef}
\vspace{0.5em}
\end{figure}

@@

We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem. In systems such as Rascal
\cite{klint2011easy}, one can see a program in a domain-specific language not just as \emph{types}
and \emph{code}, but also as \emph{data}, or an instance of a metamodel (a static type) that
defines the AST of the domain-specific language. The program in a domain-specific language
that in turn determines (defines/implies/induces) a run-time schema: the run-time structures
of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as
the structure of the run-time state.

In the absence of live programming, the run-time \emph{code} structures do not typically
change while using an application and the run-time \emph{state} changes constantly.

Live programming, however, requires reconciling changes to the program with the run-time structures
of both the code and the state. This typically means that at a certain point (a quiescent point)
during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.

\subsection{Example}
Consider the example of a simple state machine language with on-entry actions, and typed global
variables (loosely inspired by the SML language of \citet{vanRozen19}). An example state machine
and an excerpt of its abstract syntax schema is shown in Figure~\ref{LST:statemachines}. Depicted
on the left, an actual statemachine modeling the opening and closing of a door, with one global
boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the
\lstinline{closed} and \lstinline{opened} states. For the sake of brevity, the abstract syntax
schema (similar, to e.g., and Ecore metamodel~\cite{EMF}) shown on the right of the figure
omits classes for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).

\noindent
\begin{adjustwidth}{-0.5em}{0em}
\begin{minipage}[t]{0.32\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
machine Doors

var isClosed: bool = true
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.32\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
state closed
  isClosed := true
  on open => opened
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.32\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
state opened
  isClosed := false
  on close => closed
\end{lstlisting}
\end{minipage}
\end{adjustwidth}

@@

In the following we assume that the state machine is executed by an interpreter that process an incoming stream of events, and traversing run-time pointers representing transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program. The structure that the interpreter operates on can be described as derived  schema from the abstract syntax scheme shown on the right of Figure~\ref{LST:statemachines}, but specialized for a particular machine, e.g., \lstinline{Doors}.
This schema will have additional fields and associations modeling the run-time data required for execution. For the \lstinline{Doors} statemachine, this means adding the current state to the \lstinline{Machine} type (required for any machine) and the machine specific data fields, namely \lstinline{isClosed}. The run-time schema for \lstinline{Doors} is then the following\footnote{It is possible to represent the global variables at run-time using an untyped table or hashmap, but for the sake of schema migration complexity, the encoding as fields in classes is more instructive.}:

\begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
class Machine$^{++}$
  states: State*
  current: State
  isClosed: bool
\end{lstlisting}

This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.

@@

\begin{figure}[t]
  \centering
\includegraphics[width=0.5\textwidth]{figures/doorsmachine.pdf}
\caption{UML object diagram of the run-time structures of the Doors statemachine (excluding code objects)}
\label{FIG:doorsRuntime}
\end{figure}

>>>>

% \begin{figure}[t]
% \begin{adjustwidth}{-1.5em}{-1.5em}
% \begin{minipage}[t]{0.26\textwidth}
% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
% class Machine
%   name: string
%   states: State*
%   vars: Decl*
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}[t]{0.28\textwidth}
% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
% class State
%   name: string
%   onEntry: Stmt*
%   transitions: Trans*
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}[t]{0.24\textwidth}
% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
% class Trans
%   event: string
%   target: State
% $~$
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}[t]{0.24\textwidth}
% \begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
% class Decl
%   name: string
%   type: Type
% $~$
% \end{lstlisting}
% \end{minipage}
% \end{adjustwidth}
% \vspace{-1em}
% \caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
% \label{lst:machinedef}
% \vspace{0.5em}
% \end{figure}

\begin{figure}[t]
\centering
\begin{minipage}{0.3\textwidth}
\includegraphics[width=\textwidth]{figures/metamodel.pdf}
\end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.3\textwidth}
  \includegraphics[width=\textwidth]{figures/doors-ast.pdf}
  \end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.3\textwidth}
  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none]
machine Doors
var isClosed: bool = true
state closed
  isClosed := true
  on open => opened
state opened
  on close => closed
\end{lstlisting}
\end{minipage}
\caption{Definition of the abstract syntax state machine language as a UML metamodel and an example instance, as object diagram and as source code}
\label{FIG:statemachineUML}
\end{figure}

@@

We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem. In the context of executable domain-specific modeling languages~\cite{}, one can see a program in a domain-specific language as an instance of a metamodel (a static type) that
defines the AST structure of the domain-specific language. The program in a domain-specific language
then in turn determines (defines/implies/induces) a run-time schema: the run-time structures
of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
Whereas the example discussed in this section is framed in terms of metamodels and models, this should be seen as an expository device: the principles and challenges apply, e.g., in more grammar-oriented settings.

In the absence of live programming, the run-time \emph{code} structures do not typically
change while using an application and the run-time \emph{state} changes constantly.
Live programming, however, requires reconciling changes to the program with the induced run-time structures
of both the code and the state. This typically means that at a certain point (a quiescent point)
during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.

\subsection{Example}
Consider the example of a simple state machine language with on-entry actions, and typed global
variables (loosely inspired by the SML language of \citet{vanRozen19}). 
The abstract syntax (as UML class diagram) is shown on the left of Figure~\ref{FIG:statemachineUML}. 
The middle of the figure shows an instance: a state machine modeling the opening and closing of a door (in UML object notation). 
The right shows a possible textual concrete syntax modeling the same state machine. 
The state machine has one global
boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the
\lstinline{closed} and \lstinline{opened} states. For the sake of brevity, the abstract syntax
model  shown on the left of the figure
omits the details for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).

% \noindent
% \begin{adjustwidth}{-0.5em}{0em}
% \begin{minipage}[t]{0.32\textwidth}
% \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
% machine Doors

% var isClosed: bool = true
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}[t]{0.32\textwidth}
% \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
% state closed
%   isClosed := true
%   on open => opened
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}[t]{0.32\textwidth}
% \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
% state opened
%   isClosed := false
%   on close => closed
% \end{lstlisting}
% \end{minipage}
% \end{adjustwidth}

@@

In the following we assume that the state machine is executed by an interpreter that processes incoming events, and traverses run-time pointers representing the transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program. The structure that the interpreter operates on can be described as a schema as well, but derived from the abstract syntax scheme shown on the left of Figure~\ref{FIG:statemachineUML}, but decorated with additional properties and associations modeling the run-time state required for execution. This is illustrated in Figure~\ref{FIG:doorsRuntime}. The left shows an augmented metamodel adding properties \lstinline{vars} (to model variable assigments) and \lstinline{visited}, to the classes \lstinline{Machine} and \lstinline{State}, respectively.

% For the \lstinline{Doors} statemachine, this means adding the current state to the \lstinline{Machine} type (required for any machine) and the machine specific data fields, namely \lstinline{isClosed}. The run-time schema for \lstinline{Doors} is then the following\footnote{It is possible to represent the global variables at run-time using an untyped table or hashmap, but for the sake of schema migration complexity, the encoding as fields in classes is more instructive.}:

% \begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
% class Machine$^{++}$
%   states: State*
%   current: State
%   isClosed: bool
% \end{lstlisting}

% This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.

@@

\begin{figure}[t]
  \centering
\begin{minipage}{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/runtime-model.pdf}
\end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.4\textwidth}  
\includegraphics[width=\textwidth]{figures/doorsmachine.pdf}
\end{minipage}
\caption{Augmented run-time metamodel for state machines and a run-time instance of the Doors state machine during execution.}
\label{FIG:doorsRuntime}
\end{figure}
